---
title: "Retail_Analytics"
author: "Quyen Di Sabino"
date: "2026-02-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Project Overview

This project presents an end-to-end retail analytics workflow built in R, using a warehouse-style transactional dataset that mirrors real-world retail operations. The analysis focuses on transforming raw sales data into business-ready insights through data validation, feature engineering, exploratory analysis, customer segmentation, product performance evaluation, and demand forecasting.

The objective is not only to apply statistical techniques, but to translate analytical results into actionable business recommendations for pricing, customer strategy, and performance optimization.

## Important Note on Data & Interpretation

This dataset was synthetically generated for demonstration and portfolio purposes. Although it mimics a realistic retail data warehouse structure, values were randomly created.

Because of this: Certain business metrics may not behave as expected (e.g., negative margins). Promotion flags (on_promo) may not vary (e.g., all set to FALSE).

These outcomes are a result of the data generation process, not errors in the analytical workflow. Readers should focus primarily on the logic of the code.


```{r Module 01: Data Ingestion & Validation}
# ============================================================
# Retail Analytics in R
# Module 01: Data Ingestion & Validation
# ============================================================

library(tidyverse)
library(lubridate)
if (!require(janitor)) install.packages("janitor")
library(janitor)

# ------------------------------------------------------------
# 1. Load raw data
# ------------------------------------------------------------
products   <- read_csv("raw_data/dim_products.csv", show_col_types = FALSE)   %>%  clean_names()
warehouses <- read_csv("raw_data/dim_warehouses.csv", show_col_types = FALSE) %>%  clean_names()
retailers  <- read_csv("raw_data/dim_retailers.csv", show_col_types = FALSE)  %>%  clean_names()
customers  <- read_csv("raw_data/dim_customers.csv", show_col_types = FALSE)  %>%  clean_names()
orders     <- read_csv("raw_data/fact_orders.csv", show_col_types = FALSE)    %>%  clean_names()
promotions <- read_csv("raw_data/dim_promotions.csv", show_col_types = FALSE) %>%  clean_names()
shippings  <- read_csv("raw_data/fact_shippings.csv", show_col_types = FALSE) %>%  clean_names()
returns    <- read_csv("raw_data/fact_returns.csv", show_col_types = FALSE)   %>%  clean_names()
calendar   <- read_csv("raw_data/dim_date.csv", show_col_types = FALSE)       %>%  clean_names()

# ------------------------------------------------------------
# 2. Basic validation checks
# ------------------------------------------------------------

# Row counts
data.frame(
  table = c("products","warehouses","retailers","customers",
            "orders","promotions","shippings","returns","calendar"),
  rows = c(
    nrow(products),
    nrow(warehouses),
    nrow(retailers),
    nrow(customers),
    nrow(orders),
    nrow(promotions),
    nrow(shippings),
    nrow(returns),
    nrow(calendar)
  )
)

# Primary key uniqueness checks
stopifnot(nrow(products)   == n_distinct(products$product_sku))
stopifnot(nrow(customers)  == n_distinct(customers$customer_id))
stopifnot(nrow(retailers)  == n_distinct(retailers$retailer_id))
stopifnot(nrow(warehouses) == n_distinct(warehouses$warehouse_id))
stopifnot(nrow(orders)     == n_distinct(orders$order_id))

# ------------------------------------------------------------
# 3. Date sanity checks
# ------------------------------------------------------------
range(orders$order_date)
range(calendar$date)

# Shipping dates must be >= order date
shippings %>%
  filter(ship_date < order_date)

# ------------------------------------------------------------
# 4. Financial sanity checks
# ------------------------------------------------------------

orders %>%
  summarise(
    avg_price = mean(product_price),
    avg_cost  = mean(product_cost),
    pct_loss_leaders = mean(product_cost > product_price)
  )

# ------------------------------------------------------------
# 5. Save cleaned objects
# ------------------------------------------------------------
dir.create("processed_data", showWarnings = FALSE)

saveRDS(
  list(
    products   = products,
    warehouses = warehouses,
    retailers  = retailers,
    customers  = customers,
    orders     = orders,
    promotions = promotions,
    shippings  = shippings,
    returns    = returns,
    calendar   = calendar
  ),
  "processed_data/retail_data.rds"
)
```

```{r Module 02: Data Cleaning & Feature Engineering}
# ============================================================
# Retail Analytics in R
# Module 02: Data Cleaning & Feature Engineering
# ============================================================

library(tidyverse)
library(lubridate)

# ------------------------------------------------------------
# 1. Load processed data
# ------------------------------------------------------------
retail_data <- readRDS("processed_data/retail_data.rds")

products   <- retail_data$products
customers  <- retail_data$customers
retailers  <- retail_data$retailers
warehouses <- retail_data$warehouses
orders     <- retail_data$orders
shippings  <- retail_data$shippings
returns    <- retail_data$returns
calendar   <- retail_data$calendar

# ------------------------------------------------------------
# 2. Aggregate returns at order-product level
# ------------------------------------------------------------
returns_summary <- returns %>%
  group_by(order_id, product_sku) %>%
  summarise(
    return_quantity = sum(return_quantity, na.rm = TRUE),
    .groups = "drop"
  )

# ------------------------------------------------------------
# 3. Build core fact table
# ------------------------------------------------------------
fact_sales <- orders %>%
  left_join(returns_summary, by = c("order_id", "product_sku")) %>%
  mutate(
    return_quantity = replace_na(return_quantity, 0),
    net_quantity = order_quantity - return_quantity
  ) %>%
  left_join(shippings %>% select(order_id, shipping_cost, delivery_status, shipping_method),by = "order_id") %>%
  left_join(products, by = "product_sku") %>%
  left_join(customers, by = "customer_id") %>%
  left_join(retailers, by = "retailer_id") %>%
  left_join(calendar, by = c("order_date" = "date"))

# ------------------------------------------------------------
# 4. Financial feature engineering
# ------------------------------------------------------------
fact_sales <- fact_sales %>%
  mutate(
    gross_revenue = product_price * order_quantity,
    net_revenue   = gross_revenue * (1 - random_discount),
    cost_of_goods = product_cost * order_quantity,
    gross_profit  = net_revenue - cost_of_goods,
    net_profit    = gross_profit - shipping_cost,
    gross_margin  = gross_profit / net_revenue,
    markup        = (product_price - product_cost) / product_cost
  )

# ------------------------------------------------------------
# 5. Data quality flags, important
# ------------------------------------------------------------
fact_sales <- fact_sales %>%
  mutate(
    loss_leader_flag = net_profit < 0,
    full_return_flag = net_quantity <= 0,
    delayed_delivery = delivery_status == "delayed"
  )

# ------------------------------------------------------------
# 6. Time-based features
# ------------------------------------------------------------
fact_sales <- fact_sales %>%
  mutate(
    year_month = floor_date(order_date, "month"),
    week       = isoweek(order_date),
    weekday    = wday(order_date, label = TRUE),
    order_age_days = as.numeric(Sys.Date() - order_date)
  )

# ------------------------------------------------------------
# 7. Final sanity checks
# ------------------------------------------------------------
summary(select(
  fact_sales,
  gross_revenue,
  net_revenue,
  net_profit,
  gross_margin,
  markup
))

mean(fact_sales$loss_leader_flag)
mean(fact_sales$full_return_flag)

# Check impossible values
fact_sales %>%
  filter(gross_margin > 1 | gross_margin < -1) %>% 
  select(c(order_id, gross_margin))

fact_sales %>%
  filter(markup > 5 | markup < -1) %>% 
  select(c(order_id, markup))

# ------------------------------------------------------------
# 8. Save analysis-ready dataset
# ------------------------------------------------------------
saveRDS(
  fact_sales,
  "processed_data/fact_sales_enriched.rds"
)
```

```{r Module 03: Exploratory Sales & Profit Analysis}
# ============================================================
# Retail Analytics in R
# Module 03: Exploratory Sales & Profit Analysis
# ============================================================

library(tidyverse)
library(lubridate)
library(scales)

# ------------------------------------------------------------
# 1. Load enriched fact table
# ------------------------------------------------------------
fact_sales <- readRDS("processed_data/fact_sales_enriched.rds")

# ------------------------------------------------------------
# 2. Overall business KPIs
# ------------------------------------------------------------
kpi_summary <- fact_sales %>%
  summarise(
    total_revenue = sum(net_revenue, na.rm = TRUE),
    total_profit  = sum(net_profit, na.rm = TRUE),
    avg_margin    = mean(gross_margin, na.rm = TRUE),
    total_orders  = n_distinct(order_id),
    total_customers = n_distinct(customer_id)
  )

kpi_summary

# ------------------------------------------------------------
# 3. Revenue & profit trends over time
# ------------------------------------------------------------
monthly_performance <- fact_sales %>%
  group_by(year_month) %>%
  summarise(
    revenue = sum(net_revenue),
    profit  = sum(net_profit),
    .groups = "drop"
  )

monthly_performance_long <- monthly_performance %>%
  pivot_longer(
    cols = c(revenue, profit),
    names_to = "metric",
    values_to = "value"
  )

ggplot(monthly_performance_long, aes(year_month, value, color = metric)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(
    title = "Monthly Revenue vs Profit",
    x = "Month",
    y = "Amount",
    color = "Metric"
  )

# ------------------------------------------------------------
# 4. Category performance
# ------------------------------------------------------------
category_performance <- fact_sales %>%
  group_by(product_category) %>%
  summarise(
    revenue = sum(net_revenue),
    profit  = sum(net_profit),
    margin  = mean(gross_margin, na.rm = TRUE),
    .groups = "drop"
  )

category_performance_long <- category_performance %>%
  pivot_longer(
    cols = c(revenue, profit),
    names_to = "metric",
    values_to = "value"
  )

ggplot(category_performance_long, aes(reorder(product_category, value), value, fill = metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Revenue vs Profit by Product Category",
    x = "Category",
    y = "Amount",
    fill = "Metric"
  )

# plot for category margin
ggplot(category_performance,
       aes(reorder(product_category, margin), margin)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Average Profit Margin by Product Category",
    x = "Product Category",
    y = "Margin"
  )

# ------------------------------------------------------------
# 5. Regional performance
# ------------------------------------------------------------
region_performance <- fact_sales %>%
  group_by(region.y) %>%   # retailer region
  summarise(
    revenue = sum(net_revenue, na.rm = TRUE),
    profit  = sum(net_profit, na.rm = TRUE),
    .groups = "drop"
  )

region_performance_long <- region_performance %>%
  pivot_longer(
    cols = c(revenue, profit),
    names_to = "metric",
    values_to = "value"
  )

ggplot(region_performance_long, aes(reorder(region.y, value), value, fill = metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Revenue vs Profit by Retailer Region",
    x = "Region",
    y = "Amount",
    fill = "Metric"
  )
# ------------------------------------------------------------
# 6. Channel analysis
# ------------------------------------------------------------
channel_performance <- fact_sales %>%
  group_by(retailer_channel) %>%
  summarise(
    revenue = sum(net_revenue, na.rm = TRUE),
    profit  = sum(net_profit, na.rm = TRUE),
    avg_shipping_cost = mean(shipping_cost, na.rm = TRUE),
    .groups = "drop"
  )

channel_performance_long <- channel_performance %>%
  pivot_longer(
    cols = c(revenue, profit),
    names_to = "metric",
    values_to = "value"
  )

ggplot(channel_performance_long, aes(reorder(retailer_channel, value), value, fill = metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Revenue vs Profit by Retailer Channel",
    x = "Channel",
    y = "Amount",
    fill = "Metric"
  )

# Shipping Cost Visualization
ggplot(channel_performance, aes(reorder(retailer_channel, avg_shipping_cost), avg_shipping_cost)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Average Shipping Cost by Channel",
    x = "Channel",
    y = "Average Shipping Cost"
  )
# ------------------------------------------------------------
# 7. Loss leader analysis
# ------------------------------------------------------------
fact_sales %>%
  filter(loss_leader_flag) %>%
  summarise(
    loss_order_pct = n() / nrow(fact_sales),
    total_loss = sum(net_profit)
  )

# Average order value (AOV) for loss leader orders vs normal orders
fact_sales %>%
  group_by(loss_leader_flag) %>%
  summarise(
    avg_order_value = mean(product_price * order_quantity, na.rm = TRUE),
    total_revenue = sum(net_revenue)
  )

# Does profit from other products in the same order offset the loss?
fact_sales %>%
  group_by(order_id) %>%
  summarise(
    order_profit = sum(net_profit),
    loss_leader_used = any(loss_leader_flag)
  ) %>%
  group_by(loss_leader_used) %>%
  summarise(mean_order_profit = mean(order_profit))
print("Orders with loss leaders do not generate positive profit -> strategy isn't working.")

# Worse 20% loss_leaders
fact_sales %>%
  filter(loss_leader_flag) %>%
  group_by(product_sku) %>%
  summarise(total_loss = sum(net_profit)) %>%
  mutate(loss_tier = ntile(total_loss, 5)) %>% 
  filter(loss_tier == 1)

# ------------------------------------------------------------
# 8. Key takeaways (printed for report)
# ------------------------------------------------------------
cat(
  "\nKey Insights:\n",
  "- Revenue and profit trends show seasonality and volatility\n",
  "- Certain categories drive revenue\n",
  "- Loss-leading transactions are present and measurable\n",
  "- Channel-level shipping costs impact profitability\n"
)
```

```{r Module 04: Customer Analytics & RFM Segmentation}
# ============================================================
# Retail Analytics in R
# Module 04: Customer Analytics & RFM Segmentation
# ============================================================

library(tidyverse)
library(lubridate)

# ------------------------------------------------------------
# 1. Load data
# ------------------------------------------------------------
fact_sales <- readRDS("processed_data/fact_sales_enriched.rds")

analysis_date <- max(fact_sales$order_date) + 1

# ------------------------------------------------------------
# 2. RFM metrics
# ------------------------------------------------------------
rfm_table <- fact_sales %>%
  group_by(customer_id) %>%
  summarise(
    recency = as.numeric(analysis_date - max(order_date)),
    frequency = n_distinct(order_id),
    monetary = sum(net_revenue),
    .groups = "drop"
  )

# ------------------------------------------------------------
# 3. RFM scoring, quintiles
# ------------------------------------------------------------
rfm_scores <- rfm_table %>%
  mutate(
    r_score = ntile(desc(recency), 5),
    f_score = ntile(frequency, 5),
    m_score = ntile(monetary, 5),
    rfm_score = paste0(r_score, f_score, m_score)
  )

# ------------------------------------------------------------
# 4. Customer segmentation
# ------------------------------------------------------------
rfm_segments <- rfm_scores %>%
  mutate(
    segment = case_when(
      r_score >= 4 & f_score >= 4 & m_score >= 4 ~ "Champions",
      r_score >= 3 & f_score >= 3 ~ "Loyal Customers",
      r_score >= 4 & f_score <= 2 ~ "Potential Loyalists",
      r_score <= 2 & f_score <= 2 ~ "At Risk",
      TRUE ~ "Others"
    )
  )
head(rfm_segments[, c("customer_id", "segment")])

# ------------------------------------------------------------
# 5. Segment performance
# ------------------------------------------------------------
rfm_segments %>%
  group_by(segment) %>%
  summarise(
    customers = n(),
    avg_revenue = mean(monetary),
    total_revenue = sum(monetary),
    revenue_share = total_revenue / sum(rfm_segments$monetary),
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue))

# ------------------------------------------------------------
# 6. Pareto (80/20) analysis
# “Roughly 80% of effects come from 20% of causes.”
# ------------------------------------------------------------
pareto <- rfm_segments %>%
  arrange(desc(monetary)) %>%
  mutate(
    cumulative_revenue = cumsum(monetary),
    cumulative_pct = cumulative_revenue / sum(monetary)
  )

pareto %>%
  filter(cumulative_pct <= 0.8) %>%
  summarise(top_customers = n())

library(ggplot2)
library(scales)

ggplot(pareto, aes(x = seq_along(customer_id), y = cumulative_pct)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  geom_vline(xintercept = which(pareto$cumulative_pct >= 0.8)[1], linetype = "dashed", color = "steelblue") +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Pareto Analysis: Customers Contribution to Revenue",
    x = "Customers (sorted by revenue)",
    y = "Cumulative Revenue %"
  ) +
  theme_minimal()

# ------------------------------------------------------------
# 7. Key insights
# ------------------------------------------------------------
cat(
  "\nCustomer Insights:\n",
  "- Revenue is highly concentrated among a small % of customers\n",
  "- Champions and Loyal Customers drive majority of sales\n",
  "- At Risk segment requires retention focus\n"
)
```

```{r Module 05: Product Profitability & Portfolio Analysis}
# ============================================================
# Retail Analytics in R
# Module 05: Product Profitability & Portfolio Analysis
# ============================================================

library(tidyverse)
library(scales)

# ------------------------------------------------------------
# 1. Load enriched fact table
# ------------------------------------------------------------
fact_sales <- readRDS("processed_data/fact_sales_enriched.rds")

# ------------------------------------------------------------
# 2. Aggregate by product
# ------------------------------------------------------------
product_perf <- fact_sales %>%
  group_by(product_sku, product_name, product_category) %>%
  summarise(
    total_revenue = sum(net_revenue, na.rm = TRUE),
    total_profit  = sum(net_profit, na.rm = TRUE),
    avg_margin    = mean(gross_margin, na.rm = TRUE),
    total_quantity = sum(net_quantity),
    loss_orders = sum(loss_leader_flag),
    .groups = "drop"
  ) %>%
  arrange(desc(total_profit))

product_perf

# ------------------------------------------------------------
# 3. Top and bottom 10 products by profit
# ------------------------------------------------------------
product_perf %>% top_n(10, total_profit)

product_perf %>% top_n(-10, total_profit)

# ------------------------------------------------------------
# 4. Revenue vs Profit scatter (identify risky products)
# ------------------------------------------------------------
ggplot(product_perf, aes(total_revenue, total_profit, color = product_category)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype="dashed", color="red") +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Revenue vs Profit by Product",
    x = "Revenue",
    y = "Profit",
    color = "Category"
  )

# ------------------------------------------------------------
# 5. Pareto analysis: 80/20 profit
# ------------------------------------------------------------
product_perf <- product_perf %>%
  arrange(desc(total_profit)) %>%
  mutate(
    cumulative_profit = cumsum(total_profit),
    cumulative_pct = cumulative_profit / sum(total_profit)
  )

pareto_products <- product_perf %>%
  filter(cumulative_pct <= 0.8)

# % of products driving 80% profit
nrow(pareto_products) / nrow(product_perf) 

# ------------------------------------------------------------
# 6. Category-level summary
# ------------------------------------------------------------
category_perf <- fact_sales %>%
  group_by(product_category) %>%
  summarise(
    revenue = sum(net_revenue),
    profit  = sum(net_profit),
    margin  = mean(gross_margin, na.rm = TRUE)
  )

# For every $1 of revenue, how much profit do we make in each category?
category_perf %>% 
  mutate(weighted_margin = profit/revenue)

# % of products in each category are actually profitable
product_perf %>%
  group_by(product_category) %>%
  summarise(
    pct_profitable_products = mean(total_profit > 0)
  )

category_perf <- category_perf %>%
  mutate(
    profit_share = profit / sum(profit),
    revenue_share = revenue / sum(revenue)
  )

category_perf
print("If a category’s profit share > revenue share -> It punches above its weight.")

# ------------------------------------------------------------
# 7. Key insights
# ------------------------------------------------------------
cat(
  "\nProduct Insights:\n",
  "- Profit is highly concentrated: a small subset of products drives the majority of total profit (Pareto 80/20)\n",
  "- High revenue does not guarantee profitability; several products show weak or negative margins\n",
  "- Unprofitable products exist and may act as loss leaders or indicate pricing and cost issues\n",
  "- Category performance varies: some categories contribute more profit despite overall margin pressure\n"
)
```

```{r Module 06: Promotions & Pricing Impact Analysis}
# ============================================================
# Retail Analytics in R
# Module 06: Promotions & Pricing Impact Analysis
# ============================================================

library(tidyverse)
library(scales)
library(lubridate)

# ------------------------------------------------------------
# 1. Load data
# ------------------------------------------------------------
fact_sales <- readRDS("processed_data/fact_sales_enriched.rds")
promotions <- readRDS("processed_data/retail_data.rds")$promotions

# ------------------------------------------------------------
# 2. Join promotion info to orders
# ------------------------------------------------------------
fact_sales <- fact_sales %>%
  left_join(
    promotions,
    by = "product_sku",
    relationship = "many-to-many"
  ) %>%
  mutate(
    # Check if the order is during a promotion
    on_promo = !is.na(discount_percentage) & order_date >= start_date & order_date <= end_date,
   
    # Calculate the discount in dollars
    promo_discount_amt = if_else(on_promo, product_price * order_quantity * discount_percentage, 0),
    random_discount_amt = product_price * order_quantity * random_discount,
    
    # Pick the **higher discount** between promo and random
    product_discount = round(pmax(promo_discount_amt, random_discount_amt), 2)
  ) %>%
  select(-start_date, -end_date, -discount_percentage, -promo_discount_amt, -random_discount_amt, -random_discount)

# ------------------------------------------------------------
# 3. Promotion and pricing impact analysis
# ------------------------------------------------------------
# Are promotions driving incremental revenue?
promo_analysis <- fact_sales %>%
  group_by(on_promo) %>%
  summarise(
    orders = n_distinct(order_id),
    revenue = sum(net_revenue, na.rm = TRUE),
    profit = sum(net_profit, na.rm = TRUE),
    avg_margin = mean(gross_margin, na.rm = TRUE),
    AOV = mean(net_revenue, na.rm = TRUE),
    .groups = "drop"
  )

promo_analysis
print("If revenue up but profit down, then you're buying sales at the expense of margin.")

# At what discount level does profit collapse?
fact_sales %>%
  mutate(discount_bucket = ntile(product_discount, 5)) %>%
  group_by(discount_bucket) %>%
  summarise(
    avg_revenue = mean(net_revenue),
    avg_profit = mean(net_profit)
  )

# Did promotions increase quantity?
fact_sales %>%
  group_by(on_promo) %>%
  summarise(
    total_quantity = sum(net_quantity),
    avg_quantity = mean(net_quantity)
  )

# Which categories benefit more from promotions?
fact_sales %>%
  group_by(product_category, on_promo) %>%
  summarise(
    total_revenue = sum(net_revenue),
    total_profit = sum(net_profit)
  )
print("Promotions may work for seasonal items but not premium goods.")

# Time-Series Outside vs During Promotion
fact_sales %>%
  group_by(year_month, on_promo) %>%
  summarise(
    total_revenue = sum(net_revenue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = year_month,
             y = total_revenue,
             color = on_promo)) +
  geom_line(linewidth = 1.2) +
  geom_smooth(se = FALSE, linewidth = 1) +
  scale_y_continuous(labels = dollar) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(
    title = "Revenue Trend: Promotion vs Non-Promotion Periods",
    x = "Month",
    y = "Total Revenue",
    color = "Promotion Status"
  ) +
  theme_minimal()

cat("- If promotion line is consistently higher-> promotion drives revenue lift.\n",
    "- If promotion spikes only temporarily → demand may be shifted forward.\n",
    "- If non-promo revenue drops during promo → possible cannibalization."
    )
```

```{r Module 07: Demand Forecasting & Trend Analysis}
# ============================================================
# Retail Analytics in R
# Module 07: Demand Forecasting & Trend Analysis
# ============================================================

if (!require(forecast)) install.packages("forecast")
if (!require(tsibble)) install.packages("tsibble")
if (!require(feasts)) install.packages("feasts")
if (!require(fable)) install.packages("fable")

library(tidyverse)
library(lubridate)
library(forecast)
library(tsibble) # for time series
library(feasts)
library(fable)

# ------------------------------------------------------------
# 1. Load enriched sales data
# ------------------------------------------------------------
fact_sales <- readRDS("processed_data/fact_sales_enriched.rds")

# ------------------------------------------------------------
# 2. Aggregate monthly sales per product
# ------------------------------------------------------------
monthly_sales <- fact_sales %>%
  mutate(year_month = yearmonth(order_date)) %>% # fable models (ETS, ARIMA, etc.) expect a regular interval index
  group_by(product_sku, product_name, year_month) %>%
  summarise(
    quantity_sold = sum(net_quantity),
    revenue = sum(net_revenue),
    .groups = "drop"
  ) %>%
  arrange(product_sku, year_month)

# ------------------------------------------------------------
# 3. Convert to time series (tsibble)
# ------------------------------------------------------------
monthly_sales_ts <- monthly_sales %>%
  as_tsibble(index = year_month, key = product_sku)

# ------------------------------------------------------------
# 4. Forecast using ETS (Exponential Smoothing)
# ------------------------------------------------------------
# Fills missing year_month rows for each product_sku, needed for ETS sees a continuous monthly series
monthly_sales_ts <- monthly_sales_ts %>%
  fill_gaps(quantity_sold = 0, revenue = 0)

ets_model <- monthly_sales_ts %>%
  model(ets = ETS(quantity_sold))

ets_fc <- forecast(ets_model, h = "3 months")  # Forecast next 3 months

# ------------------------------------------------------------
# 5. Visualize forecast for top 5 products by revenue
# ------------------------------------------------------------
top_products <- monthly_sales %>%
  group_by(product_sku, product_name) %>%
  summarise(total_revenue = sum(revenue), .groups = "drop") %>%
  arrange(desc(total_revenue)) %>%
  slice(1:5)

ets_fc %>%
  filter(product_sku %in% top_products$product_sku) %>%
  autoplot(monthly_sales_ts, level = NULL) +
  labs(
    title = "Monthly Demand Forecast for Top 5 Products",
    x = "Month",
    y = "Quantity Sold"
  )

# ------------------------------------------------------------
# 6. Evaluate forecast accuracy (backtesting)
# ------------------------------------------------------------
# Simple train/test split for ETS evaluation
train <- monthly_sales_ts %>% filter(year_month <= yearmonth("2024 Jun"))
test  <- monthly_sales_ts %>% filter(year_month > yearmonth("2024 Jun"))

# Align products between train & test
valid_products <- intersect(
  unique(train$product_sku),
  unique(test$product_sku)
)

train_bt <- train %>% filter(product_sku %in% valid_products)
test_bt  <- test  %>% filter(product_sku %in% valid_products)

ets_bt <- train_bt %>%
  model(ets = ETS(quantity_sold))

ets_bt_fc <- forecast(ets_bt, new_data = test_bt)

accuracy(ets_bt_fc, test_bt)

# ------------------------------------------------------------
# 7. Key insights
# ------------------------------------------------------------
cat(
  "\nTime Series Forecasting Insights:\n",
  "- Monthly demand patterns were successfully modeled at the product level\n",
  "- Exponential Smoothing (ETS) performed well for stable, high-volume products.\n",
  "- Forecasting readiness depends on data continuity and product maturity.\n"
)

# ------------------------------------------------------------
# 8. Forecasting with ARIMA
# ------------------------------------------------------------
arima_model <- monthly_sales_ts %>%
  model(arima = ARIMA(quantity_sold))

arima_fc <- forecast(arima_model, h = "3 months")

arima_fc %>%
  filter(product_sku %in% top_products$product_sku) %>%
  autoplot(monthly_sales_ts, level = NULL) +
  labs(
    title = "ARIMA Forecast vs Actual Demand",
    x = "Month",
    y = "Quantity Sold"
  )

# ------------------------------------------------------------
# 9. Backtesting
# ------------------------------------------------------------

arima_bt <- train_bt %>%
  model(arima = ARIMA(quantity_sold))

arima_bt_fc <- forecast(arima_bt, new_data = test_bt)

accuracy(arima_bt_fc, test_bt)

# ------------------------------------------------------------
# 10. Choose the Final Forecasting Model (ETS vs ARIMA)
# ------------------------------------------------------------
# Fit BOTH models on the same data
models <- train_bt %>%
  model(
    ets   = ETS(quantity_sold),
    arima = ARIMA(quantity_sold)
  )

# Forecast both models
fc <- forecast(models, new_data = test_bt)

# Compare ETS vs ARIMA
acc <- accuracy(fc, test_bt)

acc %>%
  group_by(.model) %>%
  summarise(
    RMSE = mean(RMSE, na.rm = TRUE),
    MAE  = mean(MAE, na.rm = TRUE),
    ME   = mean(ME, na.rm = TRUE)
  )
```

```{r Module 08: Executive Insights & Recommendations}
# ============================================================
# Retail Analytics in R
# Module 08: Executive Insights & Recommendations
# ============================================================

library(tidyverse)

# ------------------------------------------------------------
# 1. Load analysis datasets
# ------------------------------------------------------------
fact_sales <- readRDS("processed_data/fact_sales_enriched.rds")

fact_sales <- fact_sales %>%
  left_join(
    promotions,
    by = "product_sku",
    relationship = "many-to-many"
  ) %>%
  mutate(
    # Check if the order is during a promotion
    on_promo = !is.na(discount_percentage) & order_date >= start_date & order_date <= end_date
  )

# ------------------------------------------------------------
# 2. Key Performance Indicator - KPIs
# ------------------------------------------------------------
kpis <- fact_sales %>%
  summarise(
    total_revenue = sum(net_revenue),
    total_profit  = sum(net_profit),
    avg_margin    = mean(gross_margin, na.rm = TRUE),
    loss_leader_pct = mean(loss_leader_flag)
  )

kpis

# ------------------------------------------------------------
# 3. Profit concentration, Pareto 80/20 (Product-level)
# ------------------------------------------------------------
product_profit <- fact_sales %>%
  group_by(product_sku, product_name) %>%
  summarise(
    profit = sum(net_profit),
    .groups = "drop"
  ) %>%
  arrange(desc(profit)) %>%
  mutate(
    cumulative_profit = cumsum(profit),
    cumulative_pct = cumulative_profit / sum(profit)
  )

top_profit_products <- product_profit %>%
  filter(cumulative_pct <= 0.8)

# ------------------------------------------------------------
# 4. Customer concentration
# ------------------------------------------------------------
customer_profit <- fact_sales %>%
  group_by(customer_id) %>%
  summarise(
    profit = sum(net_profit),
    .groups = "drop"
  ) %>%
  arrange(desc(profit)) %>%
  mutate(
    cumulative_profit = cumsum(profit),
    cumulative_pct = cumulative_profit / sum(profit)
  )

top_customers <- customer_profit %>%
  filter(cumulative_pct <= 0.8)


# ------------------------------------------------------------
# 5. Promotion effectiveness snapshot
# ------------------------------------------------------------
promo_compare <- fact_sales %>%
  group_by(on_promo) %>%
  summarise(
    avg_revenue = mean(net_revenue),
    avg_profit  = mean(net_profit),
    avg_margin  = mean(gross_margin, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = on_promo,
    values_from = c(avg_revenue, avg_profit, avg_margin)
  )

# Error may occur due to data was created randomly
# promo_lift <- promo_compare %>%
#    mutate(
#       revenue_lift = (avg_revenue_TRUE - avg_revenue_FALSE) /avg_revenue_FALSE,
#       profit_lift = (avg_profit_TRUE - avg_profit_FALSE) / avg_profit_FALSE,
#       margin_change = (avg_margin_TRUE - avg_margin_FALSE) / avg_margin_FALSE
#   )

# promo_lift %>%
#   select(revenue_lift_pct, profit_lift_pct, margin_change_pp) %>%
#   mutate(across(everything(), round, 2))

# ------------------------------------------------------------
# 6. Operational risk indicators
# ------------------------------------------------------------
ops_risk <- fact_sales %>%
  summarise(
    delayed_delivery_pct = mean(delayed_delivery),
    return_rate = sum(return_quantity) / sum(order_quantity)
  )

ops_risk

# Check Profit Impact of Delays
delay_impact <- fact_sales %>%
  group_by(delayed_delivery) %>%
  summarise(
    avg_profit = mean(net_profit),
    avg_margin = mean(gross_margin, na.rm = TRUE),
    .groups = "drop"
  )

# Check Quantify Financial Impact of Returns
returns_impact <- fact_sales %>%
  summarise(
    total_return_loss = sum(return_quantity * product_price),
    return_rate = sum(return_quantity) / sum(order_quantity)
  )

# ------------------------------------------------------------
# 7. Executive recommendations (printed)
# ------------------------------------------------------------
cat(
  "\nEXECUTIVE SUMMARY & RECOMMENDATIONS\n",
  "----------------------------------\n",
  "1. Revenue & Profit Concentration:\n",
  "   - A small subset of products and customers drive ~80% of profit.\n",
  "   - Focus retention and inventory planning on these segments.\n\n",
  
  "2. Product Strategy:\n",
  "   - Loss-leading products should be reviewed for repricing or discontinuation.\n",
  "   - High-revenue, low-margin products present margin optimization opportunities.\n\n",
  
  "3. Customer Strategy:\n",
  "   - Prioritize Champions & Loyal Customers with targeted promotions.\n",
  "   - Develop retention campaigns for At-Risk customers.\n\n",
  
  "4. Promotions & Pricing:\n",
  "   - Promotions may increase volume but must be controlled to protect margin.\n",
  "   - Moderate discounts outperform deep discounts in profitability.\n\n",
  
  "5. Operations:\n",
  "   - Delivery delays and returns impact profitability.\n",
  "   - Address logistics inefficiencies to improve customer experience and margins.\n"
)`{r}

```
